---
title: "Figure 4A and Extended Data Figure 5B: genome-scale modifier screens of XPR1 and IP6K inhibition"
author: "Daniel Bondeson"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: hide
    toc: true
    toc_depth: 3
---

This document regenerates the genome-scale modifier screen panels (Figure 4A and Extended
Data Figure 5B) from the Figshare dataset, writes a source-data workbook, and prints each
recomputed number next to the value in the paper. See `README.md`.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, error = FALSE,
                      fig.path = file.path(tempdir(), "fig4-"))   # keep knit chunk images out of the repo
```

# Setup

```{r packages}
library(tidyverse)
library(here)
library(scales)
library(ggrepel)
library(scattermore)
library(reshape2)    # melt() for the correlation heatmap
library(janitor)
library(writexl)

dir.create(here("results"), showWarnings = FALSE)

# Input files come from the Figshare dataset (see data/README.md). They are read from
# data/ unless FIG4_DATA_DIR points elsewhere.
data_dir <- Sys.getenv("FIG4_DATA_DIR", unset = here("data"))
data_path <- function(x) file.path(data_dir, x)

publication_theme <- theme_minimal(base_size = 11) +
  theme(
    panel.grid.minor = element_blank(),
    plot.title   = element_text(face = "bold", size = 12),
    axis.title   = element_text(size = 10),
    axis.text    = element_text(size = 9),
    strip.text   = element_text(face = "bold", size = 10),
    legend.title = element_text(face = "bold", size = 9),
    legend.text  = element_text(size = 8)
  )

drug_colors <- c("RBD27" = "#d94801", "SC919" = "darkcyan")

save_fig <- function(plot, file, width, height) {
  ggsave(here("results", file), plot = plot, width = width, height = height, units = "in")
  invisible(plot)
}

# Comparison table: every recomputed number next to its published value
checks <- list()
add_check <- function(panel, quantity, recomputed, expected, source, tol = NA, exact = FALSE) {
  ok <- if (exact) identical(as.character(recomputed), as.character(expected))
        else if (is.na(tol)) isTRUE(all.equal(recomputed, expected, tolerance = 1e-3))
        else abs(recomputed - expected) <= tol
  checks[[length(checks) + 1]] <<- tibble(
    panel = panel, quantity = quantity,
    recomputed = as.character(if (is.numeric(recomputed)) signif(recomputed, 5) else recomputed), expected = as.character(expected),
    source = source, match = ok)
}

is_control_gene <- function(g) str_detect(g, "^(NO_SITE|ONE_INTERGENIC_SITE)")
```

# Modifier screens (Fig 4A, ED Fig 5B)

Humagne enCas12a screens in RMGI cells, two library sets (C and D), 14 days of treatment
with vehicle, 200 nM RBD27 or 350 nM SC919. The input is one table of summed read counts
per set (constructs x samples); see the Figshare README for how each column was assembled.

```{r screen_data}
count_tables <- list(
  C = read_csv(data_path("RMGI_Humagne_SetC_summed_read_counts.csv"), show_col_types = FALSE),
  D = read_csv(data_path("RMGI_Humagne_SetD_summed_read_counts.csv"), show_col_types = FALSE)
)
sample_key <- read_csv(data_path("RMGI_Humagne_sample_key.csv"), show_col_types = FALSE)

# Library design: every construct scored by the portal and the gene it targets
chip <- read_csv(data_path("Humagne_construct_to_gene_map.csv"), show_col_types = FALSE)

rmgi_chronos <- read_csv(data_path("DepMap_25Q3_RMGI_CRISPR_gene_effect.csv"),
                         show_col_types = FALSE) %>%
  select(gene, chronos = chronos_gene_effect)

sapply(count_tables, function(t) c(constructs = nrow(t), samples = ncol(t) - 4))
```

## Construct-level log2 fold change against DMSO

Reads are converted to log2 reads per million within each sample (pseudocount 1, total over
every construct of the sample), replicates are collapsed by their median, and the fold
change is the drug minus the DMSO value, separately in each library set. No construct is
filtered out here: the scoring below ranks all 40,376 constructs of the library together.

```{r construct_lfc}
log2rpm <- function(x) log2(x / sum(x, na.rm = TRUE) * 1e6 + 1)

construct_lfc <- function(tab) {
  samples <- setdiff(names(tab), c("guide_id", "gene", "construct_sequence", "construct_ids"))
  rpm <- tab %>% mutate(across(all_of(samples), log2rpm))
  condition_median <- function(prefix) {
    cols <- grep(paste0("^", prefix, "_"), samples, value = TRUE)
    apply(as.matrix(rpm[cols]), 1, median, na.rm = TRUE)
  }
  tibble(guide_id = tab$guide_id, construct_sequence = tab$construct_sequence,
         DMSO = condition_median("DMSO"), RBD27 = condition_median("RBD27"),
         SC919 = condition_median("SC919")) %>%
    mutate(RBD27_vs_DMSO = RBD27 - DMSO, SC919_vs_DMSO = SC919 - DMSO)
}

# The two sets share no construct sequence. The tables also hold 334 constructs (200 in set C,
# 134 in set D, mostly with very few reads) that are not in the library design file; like the
# portal run, the scoring is restricted to the 40,376 constructs of the design file.
construct_scores <- bind_rows(C = construct_lfc(count_tables$C),
                              D = construct_lfc(count_tables$D), .id = "humagne_set") %>%
  distinct(construct_sequence, .keep_all = TRUE)

lfc <- chip %>% left_join(construct_scores, by = "construct_sequence")
stopifnot(nrow(lfc) == 40376, !anyNA(lfc$RBD27_vs_DMSO), !anyNA(lfc$SC919_vs_DMSO))
```

## Gene scores by the hypergeometric method

The gene scores come from the Broad GPP portal's hypergeometric analysis. It is
reimplemented here and checked against the portal's own output files. All constructs are
ranked by fold change. For a gene with n constructs, its k-th best construct at rank r gets
p = P(exactly k of the gene's n constructs fall among the r lowest-ranked constructs), the
hypergeometric probability with N = 40,376. This is done once from the depleted end and once
from the enriched end. The gene score is the larger of the two mean -log10(p), and the
gene's average LFC is the mean of its construct LFCs.

```{r gpp_scoring}
hypergeometric_gene_scores <- function(lfc_values, genes) {
  d <- tibble(gene = genes, lfc = lfc_values)
  N <- nrow(d)
  d <- d %>%
    mutate(asc = rank(lfc, ties.method = "min"), desc = N + 1 - asc) %>%
    group_by(gene) %>%
    mutate(n = n(),
           k_asc  = rank(asc,  ties.method = "first"),
           k_desc = rank(desc, ties.method = "first")) %>%
    ungroup() %>%
    mutate(nlp_depleted = -log10(dhyper(k_asc,  asc,  N - asc,  n)),
           nlp_enriched = -log10(dhyper(k_desc, desc, N - desc, n)))
  d %>%
    group_by(gene) %>%
    summarise(n_constructs = first(n), avg_lfc = mean(lfc),
              mean_nlp_depleted = mean(nlp_depleted), mean_nlp_enriched = mean(nlp_enriched),
              .groups = "drop") %>%
    mutate(avg_neglog10_p = pmax(mean_nlp_depleted, mean_nlp_enriched),
           direction = if_else(mean_nlp_depleted >= mean_nlp_enriched, "depleted", "enriched"))
}

gene_scores <- map(c(RBD27 = "RBD27_vs_DMSO", SC919 = "SC919_vs_DMSO"),
                   ~ hypergeometric_gene_scores(lfc[[.x]], lfc$gene))

# Compare with the portal output for every gene that has the same constructs in both
read_portal <- function(drug) {
  read_tsv(data_path(sprintf("GPP_portal_hypergeometric_%s_vs_DMSO.txt", drug)),
           show_col_types = FALSE) %>%
    clean_names() %>%
    select(gene = gene_symbol, portal_lfc = average_lfc,
           portal_nlp = average_log_p_values, portal_n = number_of_perturbations)
}
portal <- map(c(RBD27 = "RBD27", SC919 = "SC919"), read_portal)

portal_check <- imap_dfr(gene_scores, function(s, drug) {
  j <- inner_join(portal[[drug]], s, by = "gene") %>%
    filter(portal_n == n_constructs, !is_control_gene(gene))
  tibble(drug = drug, genes_compared = nrow(j),
         max_diff_average_LFC = max(abs(j$portal_lfc - j$avg_lfc)),
         max_diff_neglog10_p  = max(abs(j$portal_nlp - j$avg_neglog10_p)))
})
portal_check
for (i in seq_len(nrow(portal_check))) {
  add_check("Fig 4A", paste(portal_check$drug[i], "average LFC vs GPP portal, max difference over",
                            portal_check$genes_compared[i], "genes"),
            portal_check$max_diff_average_LFC[i], 0, "GPP portal output", tol = 1e-6)
  add_check("Fig 4A", paste(portal_check$drug[i], "average -log10 p vs GPP portal, max difference"),
            portal_check$max_diff_neglog10_p[i], 0, "GPP portal output", tol = 1e-6)
}
```

The portal also scores 599 control pseudo-genes (`NO_SITE_n`, `ONE_INTERGENIC_SITE_n`) by
pairing control constructs two at a time; the pairing in the library design file differs
from the portal run, so those entries are not reproduced and are left out of the figure.

## Fig 4A: RBD27 against SC919

```{r fig4a, fig.height=3, fig.width=4}
modifier <- gene_scores$RBD27 %>%
  select(gene, rbd_lfc = avg_lfc, rbd_p = avg_neglog10_p) %>%
  full_join(gene_scores$SC919 %>% select(gene, sc919_lfc = avg_lfc, sc919_p = avg_neglog10_p),
            by = "gene") %>%
  left_join(rmgi_chronos, by = "gene") %>%
  filter(!is_control_gene(gene))

highlight_genes <- c("NUDT4", "NUDT3", "SLC34A2", "PAX8", "KIDINS220", "SLC20A1", "SLC20A2",
                     "XPR1", "IP6K1", "IP6K2", "PPIP5K1", "PPIP5K2")

plot_df <- modifier %>% mutate(lfc_magnitude = sqrt(rbd_lfc^2 + sc919_lfc^2)) %>% arrange(lfc_magnitude)
hl <- plot_df %>% filter(gene %in% highlight_genes)

fig4a <- ggplot(plot_df, aes(rbd_lfc, sc919_lfc)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey60") +
  geom_vline(xintercept = 0, color = "grey80", linewidth = 0.3) +
  geom_hline(yintercept = 0, color = "grey80", linewidth = 0.3) +
  geom_scattermore(aes(color = chronos, alpha = lfc_magnitude),
                   pointsize = 10, pixels = c(1024, 1024)) +
  geom_point(data = hl, aes(fill = chronos), shape = 21, color = "black",
             size = 3.5, stroke = 1) +
  geom_text_repel(data = hl, aes(label = gene), size = 3.5, fontface = "bold.italic",
                  box.padding = 0.6, point.padding = 0.3, min.segment.length = 0, seed = 42,
                  max.overlaps = Inf) +
  scale_color_gradient(high = "grey80", low = "firebrick", limits = c(-1, 0),
                       oob = squish, name = "RMGI\nDependency") +
  scale_fill_gradient(high = "grey80", low = "firebrick", limits = c(-1, 0),
                      oob = squish, guide = "none") +
  scale_alpha_continuous(guide = "none") +
  coord_fixed() +
  labs(x = "RBD27 average LFC", y = "SC919 average LFC") +
  publication_theme
fig4a
save_fig(fig4a, "fig4a_modifier_screen_RBD27_vs_SC919.pdf", 4, 3)

pearson <- function(a, b) cor(a, b, use = "pairwise.complete.obs")
add_check("Fig 4A", "Pearson r, RMGI Chronos vs RBD27 average LFC",
          pearson(modifier$chronos, modifier$rbd_lfc), -0.100, "Fig 4A panel", tol = 0.001)
add_check("Fig 4A", "Pearson r, RMGI Chronos vs SC919 average LFC",
          pearson(modifier$chronos, modifier$sc919_lfc), -0.037, "Fig 4A panel", tol = 0.001)
add_check("Fig 4A", "Pearson r, RBD27 vs SC919 average LFC",
          pearson(modifier$rbd_lfc, modifier$sc919_lfc), 0.295, "Fig 4A panel", tol = 0.001)
```

## ED Fig 5B: concordance of the screen samples

Per set, arrays with fewer than 5 reads in the Presort sample are dropped, reads are
normalized within each sample over the remaining arrays, and each gene's log2 RPM is the
median over its arrays. The panel correlates each sample's gene-level log2 fold change
against Presort with the others and with the RMGI gene dependency.

```{r ed5b, fig.height=4.5, fig.width=4.5}
MIN_EARLY_READS <- 5

gene_l2fc_vs_presort <- imap_dfr(count_tables, function(tab, set_name) {
  keep <- tab %>% filter(!is.na(Presort), Presort >= MIN_EARLY_READS,
                         !is_control_gene(gene))
  samples <- setdiff(names(keep), c("guide_id", "gene", "construct_sequence", "construct_ids"))
  keep %>%
    mutate(across(all_of(samples), log2rpm)) %>%
    pivot_longer(all_of(samples), names_to = "sample", values_to = "log2rpm") %>%
    group_by(gene, sample) %>%
    summarise(log2rpm = median(log2rpm, na.rm = TRUE), .groups = "drop") %>%
    pivot_wider(names_from = sample, values_from = log2rpm) %>%
    mutate(across(-c(gene, Presort), ~ .x - Presort)) %>%
    select(-Presort) %>%
    rename_with(~ paste(set_name, .x, sep = "_"), -gene)
})
# one row per gene and set: combine the two sets side by side
gene_l2fc_wide <- gene_l2fc_vs_presort %>%
  pivot_longer(-gene, names_to = "sample_id", values_to = "L2FC", values_drop_na = TRUE) %>%
  pivot_wider(names_from = sample_id, values_from = L2FC, values_fn = median) %>%
  left_join(rmgi_chronos %>% rename(RMGI_Chronos = chronos), by = "gene")

sample_levels <- c(sort(grep("^C_", names(gene_l2fc_wide), value = TRUE)),
                   sort(grep("^D_", names(gene_l2fc_wide), value = TRUE)), "RMGI_Chronos")
cor_screen <- cor(gene_l2fc_wide[sample_levels], use = "pairwise.complete.obs")
cor_screen[upper.tri(cor_screen)] <- NA

ed5b <- melt(cor_screen, na.rm = TRUE) %>%
  mutate(Var1 = factor(Var1, levels = rev(sample_levels)), Var2 = factor(Var2, levels = sample_levels)) %>%
  ggplot(aes(Var2, Var1, fill = value)) +
  geom_tile(color = "white") +
  geom_text(aes(label = sprintf("%.2f", value)), size = 2.4) +
  scale_fill_gradient2(low = "#084594", mid = "white", high = "#d94801", midpoint = 0,
                       limits = c(-1, 1), name = "Pearson\nCorrelation") +
  publication_theme +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1), axis.title = element_blank(),
        panel.border = element_blank(), panel.background = element_blank(), aspect.ratio = 1) +
  labs(title = "Screen L2FC and gene dependency")
ed5b
save_fig(ed5b, "ed5b_screen_sample_concordance.pdf", 4.5, 4.5)
```


# Source data workbook

```{r workbook}
sheet_guide <- tribble(
  ~sheet, ~contents,
  "Counts_SetC, Counts_SetD", "Summed read counts, constructs x samples (the input of the screen analysis)",
  "Sample_key", "How each count column was assembled: source, wells, total reads",
  "Fig4A_genes", "Gene-level average LFC and -log10 p for RBD27 and SC919, RMGI Chronos",
  "Fig4A_constructs", "Construct-level LFC against DMSO for all 40,376 constructs",
  "Fig4A_portal_check", "Recomputed scores against the GPP portal output",
  "ED5B_data, ED5B_correlation", "Gene-level L2FC against Presort per sample; Pearson correlations",
  "checks_vs_paper", "Every recomputed number next to its published value")

checks_tbl <- bind_rows(checks)
write_xlsx(
  list(
    README = sheet_guide,
    Counts_SetC = count_tables$C, Counts_SetD = count_tables$D, Sample_key = sample_key,
    Fig4A_genes = modifier %>% arrange(desc(pmax(rbd_p, sc919_p, na.rm = TRUE))),
    Fig4A_constructs = lfc %>% select(construct_sequence, gene, humagne_set, guide_id,
                                      DMSO_log2rpm = DMSO, RBD27_log2rpm = RBD27, SC919_log2rpm = SC919,
                                      RBD27_vs_DMSO, SC919_vs_DMSO),
    Fig4A_portal_check = portal_check,
    ED5B_data = gene_l2fc_wide,
    ED5B_correlation = as.data.frame(cor_screen) %>% rownames_to_column("sample"),
    checks_vs_paper = checks_tbl
  ),
  here("results", "Figure4_modifier_screen_source_data.xlsx")
)
write_csv(checks_tbl, here("results", "checks_vs_paper.csv"))
knitr::kable(checks_tbl)
```

```{r session}
sessionInfo()
```
