---
title: "Figure 2 and Extended Data Figure 3: PRISM chemogenetic profiling of XPR1 and IP6K inhibitors"
author: "Daniel Bondeson"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: hide
    toc: true
    toc_depth: 3
---

This document regenerates the PRISM panels of Figure 2 (2D-F) and Extended Data Figure 3
(3B-F) from the Figshare dataset, writes a source-data workbook, and prints each recomputed
number next to its published value. See `README.md`.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, error = FALSE,
                      fig.path = file.path(tempdir(), "fig2-"))   # keep knit chunk images out of the repo
```

# Setup

```{r packages}
library(tidyverse)
library(here)
library(scales)
library(ggrepel)
library(ggrastr)    # rasterised point clouds in Fig 2F
library(ggdist)     # half violins in Fig 2E
library(patchwork)
library(broom)
library(writexl)

dir.create(here("results"), showWarnings = FALSE)

# The DepMap tables have one column per gene, so a single line is several hundred kilobytes
Sys.setenv(VROOM_CONNECTION_SIZE = 2^26)

# Input files come from the Figshare dataset (see data/README.md). They are read from
# data/ unless FIG2_DATA_DIR points elsewhere.
data_dir <- Sys.getenv("FIG2_DATA_DIR", unset = here("data"))
data_path <- function(...) file.path(data_dir, ...)

publication_theme <- theme_minimal(base_size = 11) +
  theme(
    panel.grid.minor = element_blank(),
    plot.title   = element_text(face = "bold", size = 12),
    axis.title   = element_text(size = 10),
    axis.text    = element_text(size = 9),
    strip.text   = element_text(face = "bold", size = 10),
    legend.title = element_text(face = "bold", size = 9),
    legend.text  = element_text(size = 8)
  )

drug_colors <- c("RBD27" = "#d94801", "SC919" = "darkcyan")

save_fig <- function(plot, file, width, height) {
  ggsave(here("results", file), plot = plot, width = width, height = height, units = "in")
  invisible(plot)
}

# Comparison table: every recomputed number next to its published value
checks <- list()
add_check <- function(panel, quantity, recomputed, expected, source, tol = NA, exact = FALSE) {
  ok <- if (exact) identical(as.character(recomputed), as.character(expected))
        else if (is.na(tol)) isTRUE(all.equal(recomputed, expected, tolerance = 1e-3))
        else abs(recomputed - expected) <= tol
  checks[[length(checks) + 1]] <<- tibble(
    panel = panel, quantity = quantity,
    recomputed = as.character(if (is.numeric(recomputed)) signif(recomputed, 5) else recomputed),
    expected = as.character(expected), source = source, match = ok)
}

# Genes of ED Fig 3B and 3E: phosphate transport, XPR1 and inositol pyrophosphate metabolism
phosphate_genes <- c("PAX8", "SLC34A1", "SLC34A2", "SLC34A3", "SLC20A1", "SLC20A2", "XPR1", "KIDINS220",
                     "IPMK", "IPPK", "IP6K1", "IP6K2", "IP6K3", "PPIP5K1", "PPIP5K2", "NUDT3", "NUDT4")
```

# From Sushi output to sensitivity

The PRISM screen (project 24_PSEQ_BOND_01) was processed with the CMAP Sushi pipeline
(<https://github.com/theprismlab/sushi>, commit `5be4ec2`, default settings; see
`sushi_inputs/config.json`). Sushi filters the barcode counts to the PRISM cell lines, normalizes
each PCR well with ten spiked-in control barcodes, and computes the log2 fold change of every
treated well against the vehicle wells. Its inputs and every intermediate table are in the
dataset. Everything below starts from those tables.

```{r sushi_tables}
read_sushi <- function(f) read_csv(data_path("sushi_outputs", f), show_col_types = FALSE)

filtered_counts <- read_sushi("filtered_counts.csv")
normalized      <- read_sushi("normalized_counts.csv")
l2fc_sushi      <- read_sushi("l2fc.csv")
collapsed       <- read_sushi("collapsed_l2fc.csv")
annotations     <- read_csv(data_path("PRISM_cell_line_annotations.csv"), show_col_types = FALSE)

c(cell_lines = n_distinct(collapsed$DepMap_ID), conditions = n_distinct(collapsed$sig_id))
```

## Check: log2 fold changes recomputed from the normalized counts

Sushi's step from normalized counts to fold change is simple, so it is repeated here as a check.
Technical replicates (three PCR wells per condition) are averaged, and each cell line's value
is divided by the median of its vehicle wells.

```{r l2fc_check}
tech_means <- normalized %>%
  filter(is.na(Name), !(trt_type %in% c("empty", "CB_only"))) %>%     # cell line barcodes only
  group_by(DepMap_ID, trt_type, cell_set, treatment, dose, dose_unit, day, bio_rep) %>%
  summarise(mean_normalized_n = mean(normalized_n), .groups = "drop")

vehicle <- tech_means %>% filter(trt_type == "negcon") %>%
  group_by(DepMap_ID, cell_set, day) %>%
  summarise(control_median_normalized_n = median(mean_normalized_n), .groups = "drop")

l2fc_here <- tech_means %>% filter(trt_type != "negcon") %>%
  inner_join(vehicle, by = c("DepMap_ID", "cell_set", "day")) %>%
  mutate(l2fc_recomputed = log2(mean_normalized_n) - log2(control_median_normalized_n))

l2fc_compare <- l2fc_sushi %>%
  select(DepMap_ID, cell_set, treatment, dose, dose_unit, day, bio_rep, l2fc_sushi = l2fc) %>%
  inner_join(l2fc_here %>% select(DepMap_ID, cell_set, treatment, dose, dose_unit, day, bio_rep, l2fc_recomputed),
             by = c("DepMap_ID", "cell_set", "treatment", "dose", "dose_unit", "day", "bio_rep"))

add_check("Sushi", "fold changes recomputed from normalized counts: max |difference| over all conditions",
          max(abs(l2fc_compare$l2fc_sushi - l2fc_compare$l2fc_recomputed)), 0, "Sushi l2fc.csv", tol = 1e-8)
add_check("Sushi", "fold changes compared (of those in l2fc.csv)", nrow(l2fc_compare), nrow(l2fc_sushi),
          "Sushi l2fc.csv", exact = TRUE)
```

## Sensitivity: fold change at each dose and area under the dose-response curve

The reported fold change of a cell line at a dose is Sushi's `median_l2fc`. The **AUC** is the mean,
over the doses of a drug, of the fold change capped at 1: `mean(min(2^l2fc, 1))`. A cell line that
is unaffected has an AUC of 1, one that is eliminated at every dose has an AUC near 0. RBD27 was
tested at three doses (220, 670, 2000 nM), SC919 at four (100, 300, 1000, 3000 nM).

```{r sensitivity}
dose_wide <- collapsed %>%
  mutate(sig = paste(treatment, dose, "nM", sep = ":")) %>%
  select(DepMap_ID, sig, median_l2fc) %>%
  pivot_wider(names_from = sig, values_from = median_l2fc)

auc_wide <- collapsed %>%
  group_by(DepMap_ID, treatment) %>%
  summarise(auc = mean(pmin(2^median_l2fc, 1)), n_doses = n(), .groups = "drop") %>%
  mutate(sig = paste0(treatment, ":auc")) %>%
  select(DepMap_ID, sig, auc) %>%
  pivot_wider(names_from = sig, values_from = auc)

prism_auc <- annotations %>%
  rename(ModelID = ModelID) %>%
  left_join(dose_wide, by = c("ModelID" = "DepMap_ID")) %>%
  left_join(auc_wide,  by = c("ModelID" = "DepMap_ID")) %>%
  select(ModelID, CCLE_name, StrippedCellLineName, lineage, disease,
         `RBD27:220:nM`, `RBD27:670:nM`, `RBD27:2000:nM`, `RBD27:auc`,
         `SC919:100:nM`, `SC919:300:nM`, `SC919:1000:nM`, `SC919:3000:nM`, `SC919:auc`)

stopifnot(nrow(prism_auc) == 475, !anyNA(prism_auc))
write_csv(prism_auc, here("results", "PRISM_sensitivity_by_cell_line.csv"))
add_check("Fig 2C", "cell lines screened", nrow(prism_auc), 475, "Fig 2C legend", exact = TRUE)
summary(prism_auc[c("RBD27:auc", "SC919:auc")])
```

## Fig ED3C: recovery of the cell lines in each PCR well

Each cell line is detected by its barcode. A line counts as missing from a well when it has no
reads there; wells with 40 or fewer reads are flagged as low (the `COUNT_THRESHOLD` of Sushi).
The collection of cell lines is split over two pools for PCR and sequencing.

```{r ed3c, fig.height=3.5, fig.width=3.4}
RECOVERY_THRESHOLD <- 40

recovery <- filtered_counts %>%
  filter(is.na(Name)) %>%                                  # cell lines, not control barcodes
  group_by(pcr_plate, pcr_well, cell_set) %>%
  summarise(total_num_cls = n(),
            not_detected = sum(n == 0),
            low_counts = sum(n > 0 & n <= RECOVERY_THRESHOLD),
            detected = sum(n > RECOVERY_THRESHOLD), .groups = "drop") %>%
  # Pool 1 is the smaller of the two pools, as on the published panel
  mutate(pool = paste0(if_else(total_num_cls == min(total_num_cls), "Pool 1", "Pool 2"),
                       ": ", total_num_cls, " cell lines"))
recovery %>% count(cell_set, total_num_cls, pool)

ed3c <- recovery %>%
  mutate(pcr_well = factor(pcr_well, levels = rev(sort(unique(pcr_well))))) %>%
  ggplot(aes(not_detected, pcr_well)) +
  geom_col(fill = "#d73027", width = 0.8) +
  scale_x_continuous(breaks = c(0, 2, 4), expand = expansion(mult = c(0, 0.1))) +
  facet_wrap(~ pool, scales = "free_y", ncol = 2) +
  labs(x = "Missing Cell Lines", y = "PCR Well") +
  theme_classic(base_size = 9) +
  theme(axis.text.y = element_text(size = 5.5), strip.text = element_text(size = 7, face = "bold"),
        strip.background = element_rect(fill = "grey90", color = NA), panel.spacing = unit(0.3, "lines"))
ed3c
save_fig(ed3c, "ed3c_cell_line_recovery.pdf", 3.4, 3.5)

add_check("ED Fig 3C", "most cell lines missing from any PCR well", max(recovery$not_detected), 5,
          "ED Fig 3C legend: 5 or fewer", tol = 0)

add_check("ED Fig 3C", "cell lines in pool 1 and pool 2", paste(sort(unique(recovery$total_num_cls)), collapse = " / "),
          "236 / 241", "ED Fig 3C panel", exact = TRUE)
```



# Molecular data: DepMap Public 25Q3

mRNA expression and CRISPR gene effect (Chronos) come from the public DepMap 25Q3 release
(<https://depmap.org>, CC BY 4.0). The dataset holds the cell lines of the PRISM screen only,
except for the table of pathway genes used in ED Fig 3B, which covers every model. Of the 475
screened lines, 472 have an expression profile and 400 have CRISPR data.

```{r depmap}
strip_entrez <- function(df) rename_with(df, ~ str_remove(.x, " \\(.*\\)"), -ModelID)

expression <- read_csv(data_path("DepMap_25Q3_mRNA_expression_log2TPMp1_PRISM_cell_lines.csv.gz"),
                       show_col_types = FALSE) %>% strip_entrez()
crispr <- read_csv(data_path("DepMap_25Q3_CRISPR_Chronos_gene_effect_PRISM_cell_lines.csv.gz"),
                   show_col_types = FALSE) %>% strip_entrez()
crispr_pathway <- read_csv(data_path("DepMap_25Q3_CRISPR_Chronos_pathway_genes_all_models.csv"),
                           show_col_types = FALSE) %>% strip_entrez()
biomarkers <- read_csv(data_path("PRISM_AUC_biomarker_correlations_DepMap25Q3_and_repurposing.csv"),
                       show_col_types = FALSE)

c(expression_lines = nrow(expression), expression_genes = ncol(expression) - 1,
  crispr_lines = nrow(crispr), crispr_genes = ncol(crispr) - 1)
```

# Fig 2D and 2E: sensitivity by SLC34A2 expression and lineage

Cell lines are divided by lineage (gynecologic = ovary/fallopian tube and uterus, against all
others) and by SLC34A2 mRNA expression (high above 4 log2(TPM+1), low otherwise). An absolute
threshold is used because most cell lines do not express SLC34A2 at all, so a median split
outside gynecologic tissue would separate no expression from no expression.

```{r slc_groups}
SLC34A2_THRESHOLD <- 4   # log2(TPM+1)

slc_df <- prism_auc %>%
  inner_join(expression %>% select(ModelID, SLC34A2), by = "ModelID") %>%
  drop_na(SLC34A2) %>%
  mutate(
    lin = if_else(lineage %in% c("Ovary/Fallopian Tube", "Uterus"), "Gyn.", "Other"),
    slc = if_else(SLC34A2 > SLC34A2_THRESHOLD, "SLC34A2-high", "SLC34A2-low"),
    # Fig 2D draws in this order so grey sits at the back and green on top
    Category = factor(paste(lin, slc), levels = c("Other SLC34A2-low", "Gyn. SLC34A2-low",
                                                  "Other SLC34A2-high", "Gyn. SLC34A2-high")),
    # Fig 2E facets in reading order
    Group = factor(paste(lin, slc, sep = "\n"),
                   levels = c("Gyn.\nSLC34A2-high", "Gyn.\nSLC34A2-low",
                              "Other\nSLC34A2-high", "Other\nSLC34A2-low"))) %>%
  arrange(Category)

# palette of the Fig 2C cartoon: green, blue, purple, grey
slc_cat_colors <- c("Gyn. SLC34A2-high" = "#32a354", "Gyn. SLC34A2-low" = "#4292c6",
                    "Other SLC34A2-high" = "#756bb0", "Other SLC34A2-low" = "#BDBDBD")

slc_counts <- table(slc_df$Category)
slc_counts
add_check("Fig 2D", "cell lines with an expression profile", nrow(slc_df), 472, "Fig 2C legend, Methods", exact = TRUE)
for (g in c("Gyn. SLC34A2-high", "Other SLC34A2-high", "Gyn. SLC34A2-low", "Other SLC34A2-low")) {
  add_check("Fig 2D", paste("cell lines,", g), unname(slc_counts[g]),
            c("Gyn. SLC34A2-high" = 16, "Other SLC34A2-high" = 12, "Gyn. SLC34A2-low" = 36, "Other SLC34A2-low" = 408)[[g]],
            "Fig 2D panel", exact = TRUE)
}
```

## Fig 2D

```{r fig2d, fig.height=3.4, fig.width=3.4}
r_2d <- cor(slc_df$`RBD27:auc`, slc_df$`SC919:auc`)
p_2d <- cor.test(slc_df$`RBD27:auc`, slc_df$`SC919:auc`)$p.value

fig_2d_legend <- tibble(
  Category = factor(c("Gyn. SLC34A2-high", "Other SLC34A2-high", "Gyn. SLC34A2-low", "Other SLC34A2-low"),
                    levels = levels(slc_df$Category)),
  label = sprintf("%s (%d)", c("Gyn", "Other", "Gyn", "Other"),
                  slc_counts[c("Gyn. SLC34A2-high", "Other SLC34A2-high", "Gyn. SLC34A2-low", "Other SLC34A2-low")]),
  x = c(0.40, 0.68, 0.40, 0.68), y = c(1.18, 1.18, 1.09, 1.09))
fig_2d_header <- tibble(x = c(0.02, 0.02), y = c(1.18, 1.09), label = c("SLC34A2 high:", "SLC34A2 low:"))

fig_2d <- ggplot(slc_df, aes(`RBD27:auc`, `SC919:auc`)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey50", linewidth = 0.6) +
  geom_point(aes(fill = Category, color = Category), shape = 21, size = 2.2, stroke = 0.25, alpha = 0.9) +
  geom_text(data = fig_2d_header, aes(x = x, y = y, label = label), inherit.aes = FALSE,
            hjust = 0, fontface = "bold", size = 2.8, color = "grey20") +
  geom_text(data = fig_2d_legend, aes(x = x, y = y, label = label, color = Category), inherit.aes = FALSE,
            hjust = 0, fontface = "bold", size = 2.8) +
  annotate("label", x = 0.03, y = 0.95, hjust = 0, vjust = 1,
           label = sprintf("R = %.2f\np < 0.001", r_2d), size = 3, fontface = "bold", fill = "white") +
  scale_fill_manual(values = slc_cat_colors) +
  scale_color_manual(values = c(slc_cat_colors[1:3], "Other SLC34A2-low" = "grey65")) +
  scale_x_continuous(breaks = seq(0, 1, 0.25), expand = c(0, 0)) +
  scale_y_continuous(breaks = seq(0, 1, 0.25), expand = c(0, 0)) +
  coord_cartesian(xlim = c(0, 1.02), ylim = c(0, 1.02), clip = "off") +
  labs(x = "RBD27 sensitivity (AUC)", y = "SC919 sensitivity (AUC)") +
  publication_theme +
  theme(legend.position = "none",
        axis.title.x = element_text(color = drug_colors["RBD27"], face = "bold"),
        axis.title.y = element_text(color = drug_colors["SC919"], face = "bold"),
        plot.margin = margin(t = 34, r = 8, b = 6, l = 6))
fig_2d
save_fig(fig_2d, "fig2d_RBD27_vs_SC919_AUC.pdf", 3.4, 3.4)

add_check("Fig 2D", "Pearson R, RBD27 against SC919 AUC", r_2d, 0.81, "Fig 2D panel; Results", tol = 0.005)
add_check("Fig 2D", "p < 0.001", p_2d < 0.001, TRUE, "Fig 2D panel", exact = TRUE)
```

## Fig 2E

Each group is compared with the reference group (other lineages, SLC34A2-low) by a two-sided
Mann-Whitney U test, separately for each drug.

```{r fig2e, fig.height=4.4, fig.width=5.2}
slc_long <- slc_df %>%
  select(ModelID, Group, lin, slc, SLC34A2, `RBD27:auc`, `SC919:auc`) %>%
  pivot_longer(c(`RBD27:auc`, `SC919:auc`), names_to = "Drug", values_to = "AUC") %>%
  mutate(Drug = str_remove(Drug, ":auc"))

SLC_REF <- "Other\nSLC34A2-low"
slc_pvals <- slc_long %>%
  filter(Group != SLC_REF) %>%
  group_by(Drug, Group) %>%
  summarise(p = wilcox.test(AUC, slc_long$AUC[slc_long$Group == SLC_REF & slc_long$Drug == first(Drug)],
                            exact = FALSE) %>% tidy() %>% pull(p.value), .groups = "drop") %>%
  mutate(lab = if_else(p < 1e-4, format(p, digits = 2, scientific = TRUE),
                       format(round(p, 4), scientific = FALSE, trim = TRUE)))

slc_summary <- slc_long %>%
  group_by(Group, Drug) %>%
  summarise(n = n(), median_AUC = round(median(AUC), 3), .groups = "drop") %>%
  left_join(slc_pvals %>% select(Group, Drug, p_vs_reference = p), by = c("Group", "Drug")) %>%
  mutate(Group = str_replace(Group, "\n", " "))
slc_summary

pv_pos <- slc_pvals %>%
  mutate(x = as.numeric(factor(Group, levels = levels(slc_df$Group))),
         y = if_else(Drug == "RBD27", -0.14, -0.24))

fig_2e <- ggplot(slc_long, aes(Group, AUC, fill = Drug, color = Drug)) +
  geom_hline(yintercept = 1, color = "black", linewidth = 0.4) +
  stat_slab(data = ~filter(.x, Drug == "RBD27"), side = "left", scale = 0.45, normalize = "groups",
            alpha = 0.55, color = "grey30", linewidth = 0.3) +
  stat_slab(data = ~filter(.x, Drug == "SC919"), side = "right", scale = 0.45, normalize = "groups",
            alpha = 0.55, color = "grey30", linewidth = 0.3) +
  geom_point(position = position_jitterdodge(jitter.width = 0.09, dodge.width = 0.32, seed = 1),
             size = 1.1, alpha = 0.75) +
  stat_summary(fun = median, geom = "crossbar", width = 0.22, linewidth = 0.35, color = "black",
               position = position_dodge(width = 0.32), show.legend = FALSE) +
  geom_text(data = pv_pos, aes(x = x, y = y, label = lab, color = Drug), inherit.aes = FALSE,
            size = 2.9, fontface = "bold") +
  annotate("text", x = 4, y = -0.14, label = "ref", color = drug_colors["RBD27"], size = 2.9, fontface = "bold") +
  annotate("text", x = 4, y = -0.24, label = "ref", color = drug_colors["SC919"], size = 2.9, fontface = "bold") +
  annotate("text", x = 0.72, y = 1.10, label = "RBD27", hjust = 0, color = drug_colors["RBD27"],
           fontface = "bold", size = 4.2) +
  annotate("text", x = 1.55, y = 1.10, label = "SC919", hjust = 0, color = drug_colors["SC919"],
           fontface = "bold", size = 4.2) +
  scale_fill_manual(values = drug_colors) + scale_color_manual(values = drug_colors) +
  scale_y_continuous(breaks = seq(0, 1, 0.25), expand = c(0, 0)) +
  scale_x_discrete(expand = expansion(add = 0.5)) +
  coord_cartesian(ylim = c(-0.30, 1.02), clip = "off") +
  labs(x = "U-test versus Other / SLC34A2-low  (p-value)", y = "Sensitivity (AUC)") +
  publication_theme +
  theme(panel.grid.major.x = element_blank(), legend.position = "none",
        axis.text.x = element_text(face = "bold", size = 9, lineheight = 0.95, margin = margin(t = 2)),
        axis.title.x = element_text(size = 9, margin = margin(t = 34)),
        plot.margin = margin(t = 22, r = 8, b = 4, l = 6))
fig_2e
save_fig(fig_2e, "fig2e_SLC34A2_stratified_violin.pdf", 5.2, 4.4)

published_p <- tribble(
  ~Drug,   ~Group,                      ~published, ~tol,
  "RBD27", "Gyn.\nSLC34A2-high",         1.1e-5,    0.05e-5,
  "RBD27", "Gyn.\nSLC34A2-low",          0.520,     0.001,
  "RBD27", "Other\nSLC34A2-high",        0.003,     0.001,
  "SC919", "Gyn.\nSLC34A2-high",         0.003,     0.001,
  "SC919", "Gyn.\nSLC34A2-low",          0.721,     0.001,
  "SC919", "Other\nSLC34A2-high",        0.016,     0.001)
for (i in seq_len(nrow(published_p))) {
  p <- slc_pvals %>% filter(Drug == published_p$Drug[i], Group == published_p$Group[i]) %>% pull(p)
  add_check("Fig 2E", paste("U-test p,", published_p$Drug[i], str_replace(published_p$Group[i], "\n", " "), "vs reference"),
            p, published_p$published[i], "Fig 2E panel", tol = published_p$tol[i])
}
```

# Fig 2F: molecular correlates of sensitivity

For every gene, the Pearson correlation between its profile across the PRISM cell lines and the
AUC of each drug, for mRNA expression and for CRISPR gene effect (genes measured in at least 100
of the cell lines). A positive correlation with expression means that higher expression goes with
lower sensitivity (higher AUC); a positive correlation with Chronos means the same for dependency.

```{r fig2f_data}
auc_df <- prism_auc %>% select(ModelID, RBD27 = `RBD27:auc`, SC919 = `SC919:auc`)

cor_vs_auc <- function(mat_df, label, auc_df, min_lines = 100) {
  m <- mat_df[!duplicated(mat_df$ModelID), ]
  j <- inner_join(auc_df, m, by = "ModelID")
  mat <- as.matrix(j[, -(1:3)])
  mat <- mat[, colSums(!is.na(mat)) >= min_lines, drop = FALSE]
  tibble(gene = colnames(mat), feature_set = label, n_lines = unname(colSums(!is.na(mat))),
         RBD27 = suppressWarnings(cor(mat, j$RBD27, use = "pairwise.complete.obs")[, 1]),
         SC919 = suppressWarnings(cor(mat, j$SC919, use = "pairwise.complete.obs")[, 1]))
}

biomarker_denovo <- bind_rows(cor_vs_auc(expression, "mRNA exp.", auc_df),
                              cor_vs_auc(crispr, "CRISPR dep.", auc_df)) %>% drop_na(RBD27, SC919)
biomarker_denovo %>% count(feature_set)
```

## Check against the PRISM biomarker output

The PRISM biomarker module, run at the Broad on the same DepMap release, returns the same
correlations but only for the top-ranked features. Every gene it lists is compared with the
recomputed value.

```{r fig2f_check}
supplied <- biomarkers %>%
  filter(feature_set %in% c("Expression", "CRISPR")) %>%
  mutate(feature_set = recode(feature_set, "Expression" = "mRNA exp.", "CRISPR" = "CRISPR dep.")) %>%
  select(gene = feature, feature_set, treatment, supplied = correlation_coef, n_supplied = n) %>%
  inner_join(biomarker_denovo %>% pivot_longer(c(RBD27, SC919), names_to = "treatment", values_to = "recomputed"),
             by = c("gene", "feature_set", "treatment"))
supplied %>% group_by(feature_set, treatment) %>%
  summarise(genes = n(), max_abs_difference = max(abs(supplied - recomputed)), .groups = "drop")

add_check("Fig 2F", "genes of the biomarker output (CRISPR and expression) found in the recomputed table", nrow(supplied),
          nrow(biomarkers %>% filter(feature_set %in% c("Expression", "CRISPR"))), "PRISM biomarker output", exact = TRUE)
add_check("Fig 2F", "recomputed vs biomarker output, max |difference in r|",
          max(abs(supplied$supplied - supplied$recomputed)), 0, "PRISM biomarker output", tol = 1e-5)
```

```{r fig2f, fig.height=3.3, fig.width=3.3}
paper_labels <- c("XPR1", "KIDINS220", "SYNJ2", "SOCS5", "PEAR1", "SLC20A1", "GNAT2", "SLC20A2", "USP53",
                  "TAX1BP3", "DRAP1", "KLF10", "S100A1", "SLC34A2")
f2f_lab <- biomarker_denovo %>% filter(gene %in% paper_labels)

fig_2f <- ggplot(biomarker_denovo, aes(RBD27, SC919)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey55", linewidth = 0.3) +
  geom_hline(yintercept = 0, color = "grey25", linewidth = 0.3) +
  geom_vline(xintercept = 0, color = "grey25", linewidth = 0.3) +
  ggrastr::rasterise(geom_point(aes(color = feature_set), size = 0.35, alpha = 0.45, stroke = 0),
                     dpi = 450, dev = "ragg") +
  geom_point(data = f2f_lab, aes(color = feature_set), size = 1.1) +
  geom_text_repel(data = f2f_lab, aes(label = gene, color = feature_set), size = 2.0, fontface = "bold",
                  box.padding = 0.3, min.segment.length = 0, segment.color = "black", segment.size = 0.22,
                  max.overlaps = Inf, show.legend = FALSE, seed = 1) +
  scale_color_manual(values = c("CRISPR dep." = "#31a354", "mRNA exp." = "#756bb0"), name = "Molecular Feature") +
  guides(color = guide_legend(override.aes = list(size = 1.5, alpha = 1))) +
  coord_fixed() +
  labs(x = "Pearson correlation, RBD27", y = "Pearson correlation, SC919") +
  publication_theme +
  theme(panel.border = element_rect(color = "black", fill = NA, linewidth = 0.4),
        axis.text = element_text(size = 7), axis.title = element_text(size = 8.5),
        legend.position = c(0.02, 0.98), legend.justification = c(0, 1),
        legend.background = element_rect(fill = "white", color = "black", linewidth = 0.25),
        legend.margin = margin(2, 4, 2, 2), legend.title = element_text(size = 6.5, face = "bold"),
        legend.text = element_text(size = 6), legend.key.size = unit(0.32, "cm"),
        axis.title.x = element_text(color = drug_colors["RBD27"], face = "bold"),
        axis.title.y = element_text(color = drug_colors["SC919"], face = "bold"))
fig_2f
save_fig(fig_2f, "fig2f_biomarker_concordance.pdf", 3.3, 3.3)

f2f_lab %>% arrange(feature_set, RBD27) %>% select(gene, feature_set, n_lines, RBD27, SC919) %>%
  mutate(across(c(RBD27, SC919), ~ round(.x, 3)))
```

# ED Fig 3D: expression and sensitivity across the cell lines

Cell lines are ordered by combined sensitivity (RBD27 AUC plus SC919 AUC, most sensitive first) and
drawn on a log-rank axis so that the most sensitive lines take most of the width. The numbers at the
right of each row are the Pearson correlations of that row with the RBD27 (orange) and SC919 (teal) AUC.

```{r ed3d, fig.height=6.5, fig.width=6}
target_genes <- c("XPR1", "SLC20A1", "SLC20A2", "SLC34A2", "SYNJ2", "PAX8")

heatmap_source <- prism_auc %>%
  inner_join(expression %>% select(ModelID, all_of(target_genes)), by = "ModelID") %>%
  drop_na(`RBD27:auc`, `SC919:auc`, all_of(target_genes))

cell_line_order <- heatmap_source %>% mutate(order = `RBD27:auc` + `SC919:auc`) %>% arrange(order) %>% pull(ModelID)
n_lines <- length(cell_line_order)

rank_df <- tibble(ModelID = factor(cell_line_order, levels = cell_line_order)) %>%
  mutate(rank = as.integer(ModelID), width = log2(rank + 1) - log2(rank), x = (log2(rank) + log2(rank + 1)) / 2)
logrank_breaks <- sort(unique(rank_df$x))
logrank_labels <- c(1, rep("", 8), 10, rep("", 89), 100, rep("", n_lines - 101), n_lines)

drug_long <- heatmap_source %>%
  select(ModelID,
         `RBD27 (220nM)` = `RBD27:220:nM`, `RBD27 (670nM)` = `RBD27:670:nM`, `RBD27 (2000nM)` = `RBD27:2000:nM`,
         `RBD27 (AUC)` = `RBD27:auc`,
         `SC919 (100nM)` = `SC919:100:nM`, `SC919 (300nM)` = `SC919:300:nM`, `SC919 (1000nM)` = `SC919:1000:nM`,
         `SC919 (3000nM)` = `SC919:3000:nM`, `SC919 (AUC)` = `SC919:auc`) %>%
  pivot_longer(-ModelID, names_to = "drug_dose", values_to = "value") %>%
  mutate(ModelID = factor(ModelID, levels = cell_line_order)) %>% left_join(rank_df, by = "ModelID")

expr_long <- heatmap_source %>% select(ModelID, all_of(target_genes)) %>%
  pivot_longer(-ModelID, names_to = "label", values_to = "expression") %>%
  mutate(ModelID = factor(ModelID, levels = cell_line_order),
         label = factor(label, levels = c("PAX8", "SYNJ2", "SLC34A2", "SLC20A2", "SLC20A1", "XPR1"))) %>%
  left_join(rank_df, by = "ModelID")

# Pearson correlation of every row with the two AUCs
row_cors <- heatmap_source %>%
  select(all_of(target_genes), `RBD27 (AUC)` = `RBD27:auc`, `SC919 (AUC)` = `SC919:auc`,
         `RBD27 (2000nM)` = `RBD27:2000:nM`, `RBD27 (670nM)` = `RBD27:670:nM`, `RBD27 (220nM)` = `RBD27:220:nM`,
         `SC919 (3000nM)` = `SC919:3000:nM`, `SC919 (1000nM)` = `SC919:1000:nM`,
         `SC919 (300nM)` = `SC919:300:nM`, `SC919 (100nM)` = `SC919:100:nM`) %>%
  as.matrix() %>%
  cor(., .[, c("RBD27 (AUC)", "SC919 (AUC)")], use = "pairwise.complete.obs") %>%
  as.data.frame() %>% rownames_to_column("row") %>%
  rename(RBD27 = `RBD27 (AUC)`, SC919 = `SC919 (AUC)`)

logrank_x <- function() list(scale_x_continuous(expand = c(0, 0), breaks = logrank_breaks, labels = logrank_labels),
                             scale_y_discrete(expand = c(0, 0)))
cbar <- function(title) guides(fill = guide_colourbar(barwidth = 0.5, barheight = unit(1.5, "cm"),
                                                      ticks.colour = "black", title = title, frame.colour = "black"))
no_x <- theme(axis.title.x = element_blank(), axis.ticks.x = element_blank(), axis.text.x = element_blank())
x_end <- log2(n_lines + 1)
cor_labels <- function(rows) {
  row_cors %>% filter(row %in% rows) %>%
    pivot_longer(c(RBD27, SC919), names_to = "Drug", values_to = "r") %>%
    mutate(x = x_end + if_else(Drug == "RBD27", 0.35, 1.45), label = sprintf("%.3f", r))
}
cor_text <- function(rows) list(
  geom_text(data = cor_labels(rows), aes(x = x, y = row, label = label, color = Drug), inherit.aes = FALSE,
            size = 2.6, hjust = 0), scale_color_manual(values = drug_colors, guide = "none"),
  coord_cartesian(clip = "off"))
# room between the heatmap and its colour bar for the correlation numbers
side_margin <- theme(plot.margin = margin(2, 4, 2, 5), legend.box.spacing = unit(52, "pt"))

p_expr <- ggplot(expr_long, aes(x, label, fill = expression, width = width)) + geom_tile() +
  scale_fill_gradient(low = "white", high = "#006d2c") + logrank_x() + cbar("Log2(TPM+1)") +
  cor_text(levels(expr_long$label)) + labs(x = "", y = "mRNA Exp.") + publication_theme + no_x + side_margin +
  theme(panel.border = element_rect(colour = "black", fill = NA, linewidth = 0.5))

auc_rows <- c("RBD27 (AUC)", "SC919 (AUC)")
p_auc <- drug_long %>% filter(str_detect(drug_dose, "AUC")) %>%
  mutate(drug_dose = factor(drug_dose, auc_rows)) %>%
  ggplot(aes(x, drug_dose, fill = value, width = width)) + geom_tile() +
  scale_fill_gradient(low = "#2166ac", high = "#f7f7f7", limits = c(0, 1)) + logrank_x() + cbar("AUC") +
  coord_cartesian(clip = "off") + labs(x = "", y = "") + publication_theme + no_x + side_margin

dose_rows <- c("RBD27 (220nM)", "RBD27 (670nM)", "RBD27 (2000nM)", "SC919 (100nM)", "SC919 (300nM)",
               "SC919 (1000nM)", "SC919 (3000nM)")
p_dose <- drug_long %>% filter(!str_detect(drug_dose, "AUC")) %>%
  mutate(drug_dose = factor(drug_dose, dose_rows)) %>%
  ggplot(aes(x, drug_dose, fill = value, width = width)) + geom_tile() +
  scale_fill_gradient2(low = "#2166ac", mid = "#f7f7f7", high = "#b2182b", midpoint = 0) + logrank_x() + cbar("Log2FC") +
  cor_text(dose_rows) + labs(x = "Cancer cell lines (Log Rank)", y = "Drug & Dose") + publication_theme + side_margin

fig_3d <- p_expr / p_auc / p_dose + plot_layout(heights = c(0.85, 0.3, 1))
fig_3d
save_fig(fig_3d, "ed3d_expression_vs_sensitivity_logrank.pdf", 6, 6.5)

row_cors %>% mutate(across(where(is.numeric), ~ round(.x, 3)))

published_cors <- tribble(
  ~row,              ~RBD27,   ~SC919,
  "XPR1",            -0.05,    -0.057,
  "SLC20A1",          0.042,    0.012,
  "SLC20A2",         -0.102,   -0.134,
  "SLC34A2",         -0.343,   -0.233,
  "SYNJ2",            0.189,    0.183,
  "PAX8",            -0.137,   -0.043,
  "SC919 (3000nM)",   0.75,     0.808,
  "SC919 (1000nM)",   0.72,     0.825,
  "SC919 (300nM)",    0.631,    0.826,
  "SC919 (100nM)",    0.367,    0.598,
  "RBD27 (2000nM)",   0.86,     0.749,
  "RBD27 (670nM)",    0.879,    0.731,
  "RBD27 (220nM)",    0.842,    0.702)
for (i in seq_len(nrow(published_cors))) for (d in c("RBD27", "SC919")) {
  add_check("ED Fig 3D", paste("Pearson r,", published_cors$row[i], "with", d, "AUC"),
            row_cors[[d]][row_cors$row == published_cors$row[i]], published_cors[[d]][i], "ED Fig 3D panel", tol = 0.0011)
}
add_check("ED Fig 3D", "cell lines in the heatmap", n_lines, 472, "ED Fig 3D axis", exact = TRUE)
```

# ED Fig 3B and 3E: gene dependency

```{r ed3b, fig.height=4, fig.width=6}
chronos_long <- crispr_pathway %>%
  pivot_longer(-ModelID, names_to = "Gene", values_to = "Chronos_Score") %>% drop_na() %>%
  mutate(Gene = factor(Gene, levels = phosphate_genes))
lines_per_gene <- chronos_long %>% count(Gene, name = "n_cell_lines")
lines_per_gene

fig_3b <- ggplot(chronos_long, aes(Gene, Chronos_Score)) +
  geom_jitter(width = 0.2, alpha = 0.3, color = "grey40", size = 1) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "black") +
  stat_summary(fun = median, geom = "crossbar", width = 0.5, color = "#b2182b", linewidth = 0.8) +
  labs(x = NULL, y = "CRISPR dependency (Chronos score)") + publication_theme +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, face = "bold", size = 11),
        panel.grid.major.x = element_blank())
fig_3b
save_fig(fig_3b, "ed3b_pathway_gene_Chronos_distributions.pdf", 6, 4)

add_check("ED Fig 3B", "cell lines for most genes", max(lines_per_gene$n_cell_lines), 1186, "ED Fig 3B legend", exact = TRUE)
add_check("ED Fig 3B", "cell lines for NUDT4", lines_per_gene$n_cell_lines[lines_per_gene$Gene == "NUDT4"], 45,
          "ED Fig 3B legend", exact = TRUE)
```

The correlations of ED Fig 3E use the PRISM cell lines that have CRISPR data (400). Below the
triangle, the median and variance of each profile are shown with a common direction: a larger
value always means more killing (AUC is inverted as 1 - AUC, Chronos is negated).

```{r ed3e, fig.height=5, fig.width=5}
ordered_features <- c(phosphate_genes, "RBD27 (AUC)", "SC919 (AUC)")

cor_matrix_data <- prism_auc %>% select(ModelID, `RBD27 (AUC)` = `RBD27:auc`, `SC919 (AUC)` = `SC919:auc`) %>%
  inner_join(crispr_pathway, by = "ModelID") %>%
  semi_join(crispr, by = "ModelID") %>%
  select(all_of(ordered_features))
add_check("ED Fig 3E", "PRISM cell lines with CRISPR data", nrow(cor_matrix_data), 400, "DepMap 25Q3", exact = TRUE)

crispr_cor <- cor(cor_matrix_data, use = "pairwise.complete.obs", method = "pearson")
crispr_cor[upper.tri(crispr_cor, diag = TRUE)] <- NA

cor_df <- crispr_cor %>% as.data.frame() %>% rownames_to_column("Feature_Y") %>%
  pivot_longer(-Feature_Y, names_to = "Feature_X", values_to = "Pearson_R") %>% drop_na(Pearson_R) %>%
  mutate(Feature_X = factor(Feature_X, levels = ordered_features), Feature_Y = factor(Feature_Y, levels = rev(ordered_features)))

p_tri <- ggplot(cor_df, aes(Feature_X, Feature_Y, fill = Pearson_R)) +
  geom_tile(color = "white", linewidth = 0.5) +
  scale_fill_gradient2(low = "#2166ac", mid = "#f7f7f7", high = "#b2182b", midpoint = 0, limits = c(-1, 1), name = "Pearson R",
                       guide = guide_colorbar(barheight = 8, barwidth = 1, frame.colour = "black", ticks.colour = "black")) +
  scale_x_discrete(drop = FALSE) + labs(x = NULL, y = NULL) + publication_theme +
  theme(axis.text.x = element_blank(), axis.text.y = element_text(face = "bold", size = 11), panel.grid = element_blank(),
        axis.ticks = element_blank(), plot.margin = margin(b = 0, unit = "pt"))

anno_df <- cor_matrix_data %>% pivot_longer(everything(), names_to = "Feature_X", values_to = "Raw_Value") %>% drop_na() %>%
  mutate(Effect_Score = if_else(str_detect(Feature_X, "AUC"), 1 - Raw_Value, -Raw_Value)) %>%
  group_by(Feature_X) %>% summarise(Median = median(Effect_Score), Variance = var(Effect_Score), .groups = "drop") %>%
  pivot_longer(-Feature_X, names_to = "Metric", values_to = "Score") %>%
  mutate(Feature_X = factor(Feature_X, levels = ordered_features), Metric = factor(Metric, levels = c("Median", "Variance")))

p_anno <- ggplot(anno_df, aes(Feature_X, Metric, fill = Score)) +
  geom_tile(color = "white", linewidth = 0.5) +
  scale_fill_gradient(low = "#f7fcf5", high = "#00441b", name = "Metric",
                      guide = guide_colorbar(barheight = 3, barwidth = 1, frame.colour = "black", ticks.colour = "black")) +
  scale_x_discrete(drop = FALSE) + labs(x = NULL, y = NULL) + publication_theme +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, face = "bold", size = 11),
        axis.text.y = element_text(face = "bold", size = 11), panel.grid = element_blank(), axis.ticks = element_blank(),
        plot.margin = margin(t = 0, unit = "pt"))

fig_3e <- p_tri / p_anno + plot_layout(heights = c(1, 0.4))
fig_3e
save_fig(fig_3e, "ed3e_dependency_vs_sensitivity.pdf", 5, 5)

cor_df %>% filter(Feature_X == "XPR1", Feature_Y == "KIDINS220") %>% select(Feature_X, Feature_Y, Pearson_R)
```

# ED Fig 3F: correlation with other PRISM compounds

The correlation of RBD27 and SC919 AUC with the AUC of the compounds of the public PRISM drug
repurposing screen, from the PRISM biomarker output (the repurposing data are not part of this
dataset). Each point is a compound measured in at least 430 of the cell lines; the five compounds
with the highest and the five with the lowest mean correlation are labelled.

```{r ed3f, fig.height=3.6, fig.width=3.6}
repurposing <- biomarkers %>%
  filter(feature_set == "Repurposing.Primary", dose_unit == "auc") %>%
  select(feature, treatment, correlation_coef) %>%
  pivot_wider(names_from = treatment, values_from = correlation_coef, names_prefix = "r_") %>%
  drop_na(r_RBD27, r_SC919) %>% mutate(avg_cor = (r_RBD27 + r_SC919) / 2)

repurposing_labels <- bind_rows(slice_max(repurposing, avg_cor, n = 5, with_ties = FALSE),
                                slice_min(repurposing, avg_cor, n = 5, with_ties = FALSE))
max_abs_r <- max(abs(c(repurposing$r_RBD27, repurposing$r_SC919)))

fig_3f <- ggplot(repurposing, aes(r_RBD27, r_SC919)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey50") +
  geom_vline(xintercept = 0, color = "grey80", linewidth = 0.3) + geom_hline(yintercept = 0, color = "grey80", linewidth = 0.3) +
  geom_point(alpha = 0.4, size = 1.5, color = "#32a354") +
  geom_point(data = repurposing_labels, size = 2, color = "#32a354") +
  geom_text_repel(data = repurposing_labels, aes(label = feature), size = 2.5, color = "grey40", max.overlaps = Inf,
                  box.padding = 0.9, point.padding = 0.3, min.segment.length = 0, segment.color = "black",
                  segment.size = 0.3, seed = 3) +
  coord_fixed() + publication_theme +
  # The panel shows how weakly the other compounds correlate with either drug (all |r| < 0.3), in
  # contrast to the correlation of 0.81 between RBD27 and SC919 themselves (Fig 2D).
  labs(title = sprintf("RBD27:SC919 correlation = %.2f", r_2d),
       subtitle = sprintf("Each point is one of %d other compounds (all |r| < 0.3)", nrow(repurposing)),
       x = "RBD27 correlation (Pearson r)", y = "SC919 correlation (Pearson r)") +
  theme(plot.title = element_text(size = 9, face = "bold.italic"), plot.subtitle = element_text(size = 8))
fig_3f
save_fig(fig_3f, "ed3f_PRISM_drug_drug_correlation.pdf", 3.6, 3.6)

published_labels <- c("DEPRODONE PROPIONATE", "CEP-32496", "PD-407824", "AZ20", "PENFLURIDOL",
                      "EDITOL", "SNDX-5613", "OCLACITINIB", "IPI549", "SAR125844")
add_check("ED Fig 3F", "the ten labelled compounds are the top five and bottom five",
          paste(sort(repurposing_labels$feature), collapse = "; "), paste(sort(published_labels), collapse = "; "),
          "ED Fig 3F panel", exact = TRUE)
add_check("ED Fig 3F", "compounds plotted", nrow(repurposing), 284, "PRISM biomarker output", exact = TRUE)
add_check("ED Fig 3F", "title: correlation of RBD27 and SC919 AUC", r_2d, 0.81, "ED Fig 3F panel title (= Fig 2D)", tol = 0.005)
add_check("ED Fig 3F", "largest |correlation| of another compound with either drug is below 0.3",
          max_abs_r < 0.3, TRUE, "ED Fig 3F panel", exact = TRUE)
```

# Source data workbook

```{r workbook}
sheet_guide <- tribble(
  ~sheet, ~contents,
  "Sensitivity_by_cell_line", "475 cell lines: lineage, fold change at each dose and AUC for RBD27 and SC919",
  "Fig2D_E_cell_lines", "472 cell lines with SLC34A2 expression, group, and AUC (Fig 2D, 2E)",
  "Fig2E_stats", "Group sizes, medians and Mann-Whitney U p-values against Other / SLC34A2-low",
  "Fig2F_correlations", "Pearson correlation of every gene's expression and Chronos score with RBD27 and SC919 AUC",
  "Fig2F_vs_biomarker_output", "Recomputed correlations next to the PRISM biomarker output",
  "ED3B_Chronos", "Chronos score of the pathway genes in every DepMap model",
  "ED3C_recovery", "Cell lines detected, low and missing in each PCR well",
  "ED3D_correlations", "Pearson correlation of each heatmap row with the AUC of each drug",
  "ED3E_correlations", "Pearson correlation matrix and median / variance track",
  "ED3F_repurposing", "Correlation of both drugs with each repurposing compound",
  "Sushi_l2fc_check", "Fold changes recomputed from normalized counts next to Sushi's",
  "checks_vs_paper", "Every recomputed number next to its published value")

checks_tbl <- bind_rows(checks)
write_xlsx(
  list(
    README = sheet_guide,
    Sensitivity_by_cell_line = prism_auc,
    Fig2D_E_cell_lines = slc_df %>% select(ModelID, CCLE_name, lineage, disease, SLC34A2_expression = SLC34A2,
                                           lineage_group = lin, SLC34A2_group = slc, `RBD27:auc`, `SC919:auc`),
    Fig2E_stats = slc_summary,
    Fig2F_correlations = biomarker_denovo,
    Fig2F_vs_biomarker_output = supplied,
    ED3B_Chronos = chronos_long,
    ED3C_recovery = recovery,
    ED3D_correlations = row_cors,
    ED3E_correlations = crispr_cor %>% as.data.frame() %>% rownames_to_column("feature"),
    ED3F_repurposing = repurposing,
    Sushi_l2fc_check = l2fc_compare,
    checks_vs_paper = checks_tbl),
  here("results", "Figure2_source_data.xlsx"))
write_csv(checks_tbl, here("results", "checks_vs_paper.csv"))
knitr::kable(checks_tbl)
```

```{r session}
sessionInfo()
```
